Objective-first design of high-efficiency, small-footprint couplers between arbitrary nanophotonic waveguide modes

### 2012-03-06 presentations/ob1_wg_paper/paper.tex
Final submission to Opec, with response to reviewers.
Updated bicg theory.

\begin{thebibliography}{99}
\bibitem{fibergrating} Y. Tang, Z. Wang, L. Wosinski, U. Westergren, and S. He,
    ``Highly efficient nonuniform grating coupler for silicon-on-insulator 
    nanophotonic circuits,''
    Opt. Lett. \textbf{35}, 1290-1292 (2010).
\bibitem{ridge}  K. K. Lee, D. R. Lim, L.C. Kimerling, J. Shin, and F. Cerrina, 
    ``Fabrication of ultralow-loss Si/SiO2 waveguides by roughness reduction,''
    Opt. Lett. \textbf{26}, 1888-1890 (2001).
\bibitem{pcslow} Y. A. Vlasov, M. O'Boyle, H. F. Hamann, and S. J. McNab,
    ``Active control of slow light on a chip with photonic crystal waveguides,''
    Nature \textbf{438}, 65-69 (2005).
\bibitem{slotfocus} M. Lipson, 
    ``Guiding, modulating, and emitting light on 
    silicon-challenges and opportunities,'' 
    J. Lightwave Technol. \textbf{23}, 4222-4238 (2005).
\bibitem{active} J. Van Campenhout, P. Rojo Romeo, P. Regreny, C. Seassal, 
    D. Van Thourhout, S. Verstuyft, L. Di Cioccio, J.-M. Fedeli, 
    C. Lagahe, and R. Baets, 
    ``Electrically pumped InP-based microdisk lasers integrated with a 
    nanophotonic silicon-on-insulator waveguide circuit,'' 
    Opt. Express \textbf{15}, 6744-6749 (2007).
\bibitem{metallic} L. Tang, S. E. Kocabas, S. Latif, A. K. Okyay, 
    D. S. Ly-Gagnon, K. C. Saraswat, and D. A. B. Miller, 
    ``Nanometre-scale germanium photodetector enhanced by a 
    near-infrared dipole antenna,'' 
    Nature Photonics \textbf{2}, 226 – 229 (2008).
% \bibitem{fwadia} V. R. Almeida, R. R. Panepucci, and M. Lipson, 
%     ``Nanotaper for compact mode conversion,'' 
%     Opt. Lett. \textbf{28}, 1302-1304 (2003) 
% \bibitem{wwadia} S. G. Johnson, P. Bienstman,  M. A. Skorobogatiy, 
%     M. Ibanescu1, E. Lidorikis, and J. D. Joannopoulos,
%     ``Adiabatic theorem and continuous coupled-mode theory for 
%     efficient taper transitions in photonic crystals,''
%     Phys. Rev. E \textbf{66}, 066608 (2002)
% \bibitem{deriv} F. Wang, J. S. Jensen, O. Sigmund, 
%     ``Robust topology optimization of photonic crystal waveguides with 
%     tailored dispersion properties.'' 
%     J. Opt. Soc. Am. B \textbf{28}, 387-397 (2011)
% \bibitem{boydbook} S. Boyd, and L. Vandenberghe, 
%     \emph{Convex Optimization} 
%     (Cambridge University Press, 2004)
\bibitem{prevwork} J. Lu, S. Boyd, and J. Vuckovic, 
    ``Inverse design of a three-dimensional nanophotonic resonator,''
    Opt. Express \textbf{19}, 10563-10.750 (2011). 
\bibitem{code} \url{www.github.com/JesseLu/objective-first}
\end{thebibliography}

@@

% How is it applied to waveguide coupler design?
\section{Objective-first design of waveguide couplers}
We now demonstrate the objective-first approach
    by designing waveguide couplers in two-dimensions.
Specifically, we work in the two-dimensional transverse electric mode,
    choosing $H_z$ as the field variable $x$,
    and $\epsilon^{-1}$ (the inverse of the permittivity) 
    as the structure variable $p$.
This results in the following form of the physics residual,
    \begin{equation}
    \|g(x,p)\|^2 = 
    \|g(H_z, \epsilon^{-1})\|^2 = 
    \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu_0 \omega^2 H_z \|^2,
    \end{equation}
    where $\omega$ is the angular frequency,
    and $\mu_0$ is the permeability of free-space.

@@

Lastly, we employ an alternating directions strategy,
    where both $x$ and $p$ are solved iteratively\cite{prevwork}
    and limit the allowable values of $\epsilon$ to be between
    the permittivity of vacuum and of silicon,
    \begin{equation}
    \epsilon_0 \le \epsilon \le \epsilon_\text{silicon}.
    \end{equation}
Note that a completely binary structure,
    where $\epsilon = \{\epsilon_0, \epsilon_\text{silicon}\}$,
    is needed for manufacturing purposes; this will be pursued in a future work.
That said, the final designs presented here 
    all have significant portions which are already binary.

\subsection{Coupler designs}
We now present five results which suggest our method
    is indeed able to design arbitrary single-mode, linear nanophotonic devices.
These results comprise of couplers between 
    the following nanophotonic waveguides:
    \begin{enumerate}
    \item coupling between waveguides of different refractive index and width
        (Fig.~\ref{fig:fiber});
    \item coupling between waveguide modes of different order and symmetry
        (Fig.~\ref{fig:mode});
    \item coupling between waveguides that confine light 
        using different principles 
        (index guided vs. distributed Bragg reflection guided), 
        i.e., between a slab waveguide and a photonic crystal fiber
        (Fig.~\ref{fig:aircore})
    \item coupling from a dielectric to a plasmonic metal-insulator-metal 
        waveguide (Fig.~\ref{fig:mim}); and
    \item coupling from a dielectric waveguide to a (plasmonic) metal wire
        (Fig.~\ref{fig:wire}).
    \end{enumerate}
All the designs presented are both highly efficient 
    ($\sim 98\%$ coupling efficiency)
    and extremely compact 
    (with footprints of only 1.5-4 square vacuum wavelengths).
In addition, good starting points were not required; 
    all initial designs were simply $\epsilon = 9$ everywhere
    (a somewhat arbitrary guess, other values work as well).
Finally, only 15 minutes on a single-core personal computer
    were needed to obtain each design.

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{1} 
    \caption{Coupler from a narrow, high-index ($\epsilon=12.25$) waveguide to
            a wide, low-index ($\epsilon=2.25$) waveguide. 
        The $H_z^\text{perfect}$ boundary values used as 
            the design objective are shown in the upper-left plots, 
            the generated structure is shown in the upper-right plot, and
            the simulated $H_z$ fields of the device are shown 
            in the bottom two plots.
        The computed efficiency of the coupler is high, $99.8\%$, and
            the device is also extremely compact, 
            convering only $36 \times 76$ grid points,
            where the vacuum wavelength is 42 grid points
            (footprint of $1.55$ square vacuum wavelengths).
        Computation time was 15 minutes on a personal computer.}
    \label{fig:fiber}
\end{figure}
\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{2} 
    \caption{Coupler that converts the fundamental waveguide mode to the
            second-order waveguide mode
            (a movie of the design progress is included as a
            supplementary material).
        This problem is quite difficult since the two modes are of 
            opposite symmetry.
        For example, adiabatic approaches cannot be applied to this case.
        However, our method produces a device 
            (which has the same dimensions and vacuum wavelength as 
            Fig.~\ref{fig:fiber})  
            with a coupling efficiency of $98.0\%$. 
        Computation time was 15 minutes on a personal computer.
        }
    \label{fig:mode}
\end{figure}
\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{3}
    \caption{Coupler between a dielectric slab waveguide to 
            an air-core waveguide.
        Here, not only are the modes of opposite symmetry,
            but the output waveguide operates on a fundamentally different
            principle (distributed reflection) than the input waveguide 
            (index guided).
        The device still achieves an efficiency of $98.9\%$, demonstrating the
            versatility of our method.
        The vacuum wavelength is 25 grid points, 
            while the device footprint is still $36 \times 76$ grid points
            (footprint of 4.38 square vacuum wavelengths).
        Computation time was 15 minutes on a personal computer.
        }
        \label{fig:aircore}
\end{figure}
\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{4}
    \caption{Coupler between a dielectric slab waveguide to 
            a plasmonic metal-insulator-metal waveguide.
        The efficiency of the device is $97.5\%$ and 
            has the same wavelength and footprint as the device in
            Fig.~\ref{fig:aircore}.
        Computation time was 15 minutes on a personal computer.
        }
        \label{fig:mim}
\end{figure}
\clearpage
\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{5}
    \caption{Coupler between a dielectric slab waveguide to 
            a plasmonic wire waveguide.
        The efficiency of the device is $99.1\%$ and 
            has the same wavelength and footprint as the device in
            Fig.~\ref{fig:aircore}.
        Computation time was 15 minutes on a personal computer.
        }
        \label{fig:wire}
\end{figure}

@@

>>>>

\begin{thebibliography}{99}
\bibitem{fibergrating} Y. Tang, Z. Wang, L. Wosinski, U. Westergren, and S. He,
    ``Highly efficient nonuniform grating coupler for silicon-on-insulator 
    nanophotonic circuits,''
    Opt. Lett. \textbf{35}, 1290-1292 (2010).
\bibitem{ridge}  K. K. Lee, D. R. Lim, L.C. Kimerling, J. Shin, and F. Cerrina, 
    ``Fabrication of ultralow-loss Si/SiO2 waveguides by roughness reduction,''
    Opt. Lett. \textbf{26}, 1888-1890 (2001).
\bibitem{pcslow} Y. A. Vlasov, M. O'Boyle, H. F. Hamann, and S. J. McNab,
    ``Active control of slow light on a chip with photonic crystal waveguides,''
    Nature \textbf{438}, 65-69 (2005).
\bibitem{slotfocus} M. Lipson, 
    ``Guiding, modulating, and emitting light on 
    silicon-challenges and opportunities,'' 
    J. Lightwave Technol. \textbf{23}, 4222-4238 (2005).
\bibitem{active} J. Van Campenhout, P. Rojo Romeo, P. Regreny, C. Seassal, 
    D. Van Thourhout, S. Verstuyft, L. Di Cioccio, J.-M. Fedeli, 
    C. Lagahe, and R. Baets, 
    ``Electrically pumped InP-based microdisk lasers integrated with a 
    nanophotonic silicon-on-insulator waveguide circuit,'' 
    Opt. Express \textbf{15}, 6744-6749 (2007).
\bibitem{metallic} L. Tang, S. E. Kocabas, S. Latif, A. K. Okyay, 
    D. S. Ly-Gagnon, K. C. Saraswat, and D. A. B. Miller, 
    ``Nanometre-scale germanium photodetector enhanced by a 
    near-infrared dipole antenna,'' 
    Nature Photonics \textbf{2}, 226-229 (2008).
% \bibitem{fwadia} V. R. Almeida, R. R. Panepucci, and M. Lipson, 
%     ``Nanotaper for compact mode conversion,'' 
%     Opt. Lett. \textbf{28}, 1302-1304 (2003) 
% \bibitem{wwadia} S. G. Johnson, P. Bienstman,  M. A. Skorobogatiy, 
%     M. Ibanescu1, E. Lidorikis, and J. D. Joannopoulos,
%     ``Adiabatic theorem and continuous coupled-mode theory for 
%     efficient taper transitions in photonic crystals,''
%     Phys. Rev. E \textbf{66}, 066608 (2002)
% \bibitem{deriv} F. Wang, J. S. Jensen, O. Sigmund, 
%     ``Robust topology optimization of photonic crystal waveguides with 
%     tailored dispersion properties.'' 
%     J. Opt. Soc. Am. B \textbf{28}, 387-397 (2011)
% \bibitem{boydbook} S. Boyd, and L. Vandenberghe, 
%     \emph{Convex Optimization} 
%     (Cambridge University Press, 2004)
\bibitem{prevwork} J. Lu, S. Boyd, and J. Vuckovic, 
    ``Inverse design of a three-dimensional nanophotonic resonator,''
    Opt. Express \textbf{19}, 10563-10750 (2011). 
\bibitem{veronis} G. Veronis, and S. Fan, 
    ``Theoretical investigations of compact couplers between dielectric slab 
    waveguides and two-dimensional metal-dielectric-metal plasmonic waveguides,'' 
    Opt. Express \textbf{15}, 1211-1221 (2007).
\bibitem{yang} R. Yang, R. A. Wahsheh, Z. Lu, and M. A. G. Abushagur, 
    ``Efficient light coupling between dielectric slot waveguide and plasmonic
    slot waveguide,'' Opt. Lett. \textbf{35}, 649-651 (2010).
\bibitem{code} \url{www.github.com/JesseLu/objective-first}
\end{thebibliography}

@@

% How is it applied to waveguide coupler design?
\section{Objective-first design of waveguide couplers}
We now demonstrate the objective-first approach
    by designing waveguide couplers in two-dimensions.
As such, we do not analyze effects which are only present in full 
    three-dimensional systems, such as out-of-plane losses.
Specifically, we work in the two-dimensional transverse electric mode,
    choosing $H_z$ as the field variable $x$,
    and $\epsilon^{-1}$ (the inverse of the permittivity) 
    as the structure variable $p$.
This results in the following form of the physics residual,
    \begin{equation}
    \|g(x,p)\|^2 = 
    \|g(H_z, \epsilon^{-1})\|^2 = 
    \| \nabla \times \epsilon^{-1} \nabla \times H_z - \mu_0 \omega^2 H_z \|^2,
    \end{equation}
    where $\omega$ is the angular frequency,
    and $\mu_0$ is the permeability of free-space.

@@

Lastly, we employ an alternating directions strategy,
    where both $x$ and $p$ are solved iteratively\cite{prevwork}
    and limit the allowable values of $\epsilon$ to be between
    the permittivity of vacuum and of silicon,
    \begin{equation}
    \epsilon_0 \le \epsilon \le \epsilon_\text{silicon}.
    \end{equation}
Note that a completely binary structure,
    where $\epsilon = \{\epsilon_0, \epsilon_\text{silicon}\}$,
    is needed for manufacturing purposes; this will be pursued in a future work
    along with an analysis of robustness to fabrication imperfections.
That said, the final designs presented here 
    all have significant portions which are already binary.

\subsection{Coupler designs}
We now present five results which suggest our method
    is indeed able to design arbitrary single-mode, linear nanophotonic devices.
These results comprise of couplers between 
    the following nanophotonic waveguides:
    \begin{enumerate}
    \item coupling between waveguides of different refractive index and width
        (Fig.~\ref{fig:fiber});
    \item coupling between waveguide modes of different order and symmetry
        (Fig.~\ref{fig:mode});
    \item coupling between waveguides that confine light 
        using different principles 
        (index guided vs. distributed Bragg reflection guided), 
        i.e., between a slab waveguide and a photonic crystal fiber
        (Fig.~\ref{fig:aircore})
    \item coupling from a dielectric to a plasmonic metal-insulator-metal 
        waveguide (Fig.~\ref{fig:mim}); and
    \item coupling from a dielectric waveguide to a (plasmonic) metal wire
        (Fig.~\ref{fig:wire}).
    \end{enumerate}
All the designs presented are both highly efficient 
    ($\sim 98\%$ coupling efficiency, in contrast to previous designs
    which achieve only $\sim 70\%$ coupling efficiency\cite{veronis,yang})
    and extremely compact 
    (with footprints of only 1.5-4 square vacuum wavelengths).

Note, however, that the small footprint of the designs presented does not 
    imply small, unresolvable feature sizes.
On the other hand, assuming a vacuum wavelength of 1550 nm,
    the smallest possible feature in any of the examples presented is still
    37 nm $\times$ 37 nm (see Fig.~\ref{fig:fiber}, for example),
    as given by the size of a single grid point;
    placing virtually all designs within reach of current-generation lithographic
    systems.

In addition, good starting points were not required; 
    all initial designs were simply $\epsilon = 9$ everywhere
    (a somewhat arbitrary guess, other values work as well).
Finally, only 15 minutes on a single-core personal computer
    were needed to obtain each design.

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{1} 
    \caption{Coupler from a narrow, high-index ($\epsilon=12.25$) waveguide to
            a wide, low-index ($\epsilon=2.25$) waveguide. 
        The $H_z^\text{perfect}$ boundary values used as 
            the design objective are shown in the upper-left plots, 
            the generated structure is shown in the upper-right plot, and
            the simulated $H_z$ fields of the device are shown 
            in the bottom two plots.
        The computed efficiency of the coupler is high, $99.8\%$, and
            the device is also extremely compact, 
            convering only $36 \times 76$ grid points,
            where the vacuum wavelength is 42 grid points
            (footprint of $1.55$ square vacuum wavelengths).
        Computation time was 15 minutes on a personal computer.}
    \label{fig:fiber}
\end{figure}
\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{2} 
    \caption{Coupler that converts the fundamental waveguide mode to the
            second-order waveguide mode.
        This problem is quite difficult since the two modes are of 
            opposite symmetry.
        For example, adiabatic approaches cannot be applied to this case.
        However, our method produces a device 
            (which has the same dimensions and vacuum wavelength as 
            Fig.~\ref{fig:fiber})  
            with a coupling efficiency of $98.0\%$. 
        Computation time was 15 minutes on a personal computer.
        }
    \label{fig:mode}
\end{figure}
\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{3}
    \caption{Coupler between a dielectric slab waveguide to 
            an air-core waveguide.
        Here, not only are the modes of opposite symmetry,
            but the output waveguide operates on a fundamentally different
            principle (distributed reflection) than the input waveguide 
            (index guided).
        The device still achieves an efficiency of $98.9\%$, demonstrating the
            versatility of our method.
        The vacuum wavelength is 25 grid points, 
            while the device footprint is still $36 \times 76$ grid points
            (footprint of 4.38 square vacuum wavelengths).
        Computation time was 15 minutes on a personal computer.
        }
        \label{fig:aircore}
\end{figure}
\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{4}
    \caption{Coupler between a dielectric slab waveguide to 
            a plasmonic metal-insulator-metal waveguide.
        The efficiency of the device is $97.5\%$ and 
            has the same wavelength and footprint as the device in
            Fig.~\ref{fig:aircore}.
        Computation time was 15 minutes on a personal computer.
        }
        \label{fig:mim}
\end{figure}
\begin{figure}[htb]
    \centering
    \includegraphics[width=0.8\textwidth]{5}
    \caption{Coupler between a dielectric slab waveguide to 
            a plasmonic wire waveguide.
        The efficiency of the device is $99.1\%$ and 
            has the same wavelength and footprint as the device in
            Fig.~\ref{fig:aircore}.
        Computation time was 15 minutes on a personal computer.
        }
        \label{fig:wire}
\end{figure}
\clearpage

@@

% \clearpage
